Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-07-08 Manuscript/manuscript.tex
Results edits
Smaller edits following from edits to methods to try to better
distinguish between the different parameterizations and simulations.

The above simulations held the mean of patch quality distribution constant, and explored the effect of increasing habitat heterogeneity around that mean (i.e. by expanding the quality distribution symmetrically about 1).  To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e. by expanding the quality distribution non-symmetrically about 1).  For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

@@

The pathogen persisted in 70$\%$ of the 10000 simulations in the baseline model with symmetric patch quality distributions, spreading through the entire susceptible host population (i.e. a pandemic) in 37$\%$ of the simulations.  Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).  The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).  The three alternative parameterizations (lattice structure, $\alpha = 0$, $\delta = 0.1$) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e. fewer pandemics), and the habitat quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).

@@

To separate the influence of high and low quality patches, we explored how susceptible and infectious patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.  Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).  Conversely, a shift towards low quality patches decreased infectious occupancy, while a shift towards high quality patches increased infectious occupancy (Fig~\ref{sens}(b)).  These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).

However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).  Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high longevity.  On the other hand, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.  The model in which the extinction rate does not scale with quality (i.e. $\alpha = 0$) produced very similar results (Fig~\ref{sens_x0}(c)).  However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevity (Fig~\ref{sens_lattice}(c)).

@@

To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectiouss only briefly and infrequently (Fig~\ref{simvis}).  In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).  This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infectious.  A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).

\begin{figure}
\centering
\includegraphics[height = 9cm]{simvis.png}
\caption{Results from a single simulation with variance 0.2 and longevity 80.  The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
\label{simvis}
\end{figure}

>>>>

The above simulations held the mean of patch quality distribution constant, and explored the effect of increasing habitat heterogeneity around that mean (i.e. by expanding the quality distribution symmetrically about 1).  To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e. by expanding the quality distribution asymmetrically about 1).  For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

@@

In the baseline parameterization, pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).  The effect of longevity, however, depended on the shape of the patch quality distribution.  In the simulations with patch quality distributions symmetric around 1, low variance metapopulations favored endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowered the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).  The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e. fewer pandemics), and the patch quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).

@@

When we explored the influence of high and low quality patches separately (i.e. in simulations that expanded the patch quality distribution asymmetrically), we found that, relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution to include low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).  Conversely, a shift towards low quality patches decreased infectious occupancy, while a shift towards high quality patches increased infectious occupancy (Fig~\ref{sens}(b)).  These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).

However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).  Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high pathogen longevity.  On the other hand, though low quality patch distributions increase total occupancy at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.  The model in which the extinction rate does not scale with quality (i.e. $\alpha = 0$) produced very similar results, though with less pronounced differences between the different quality distributions (Fig~\ref{sens_x0}(c)).  However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevities (Fig~\ref{sens_lattice}(c)).

@@

To understand the mechanisms behind these effects, we further examined within patch dynamics in the baseline model.  To compare across the full range of patch qualities, and to ensure the opportunity for both suscepible and infectious occupancy, we limit this analysis to high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).  In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).  This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infectious.  A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).

\begin{figure}
\centering
\includegraphics[height = 9cm]{simvis.png}
\caption{Results from a single representative simulation with variance 0.2 and longevity 80.  The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
\label{simvis}
\end{figure}
